Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-15 arxiv-version/main.tex
Add multiclass perceptron

% Math symbols and commands
\newcommand{\R}{\mathbb{R}}  % set of real numbers
\newcommand{\indicator}[1]{\mathbf{1}\left[#1 \right]} % indicator
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle} % inner product
\newcommand{\norm}[1]{\left\| #1 \right\|}  % norm of a vector or a matrix

@@

We study the online multiclass classification problem with bandit feedback. The
protocol of the problem is stated below. The problem was initially proposed by
\cite{Kakade-Shalev-Shwartz-Tewari-2008}. The corresponding problem with full
information feedback is solved by the classical \textsc{Multiclass Perceptron}
algorithm and its variants (see e.g.
\cite{Crammer-Dekel-Keshet-Shalev-Shwartz-Singer-2006}). The problem is quite
natural and potential applications include online advertising and online content
optimization where the feedback corresponds to a user's click.

\begin{algorithm}[h]
\caption{Online multiclass classification with bandit feedback
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$}
\FOR{$t=1,2,\dots$}
\STATE{Receive feature vector $x_t$}
\STATE{Environment \emph{secretly} chooses correct class label $y_t \in \{1,2,\dots,K\}$}
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Receive feedback $z_t \in \{0,1\}$ where $z_t = \begin{cases}
1 & \text{if $\widehat y_t = y_t$,} \\
0 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

In this paper, we focus on the special case when the examples are \emph{linearly
separable with a margin}. In the full-information feedback setting, it is well
know that if the examples have norm at most $R$ and are linearly separable with
a margin $\gamma$ then the \textsc{Multiclass Perceptron} algorithm makes at
most $(R/\gamma)^2$ mistakes. This result is very satisfactory since the upper
bound on the number of mistakes is information-theoretically optimal and at the
same time the \textsc{Multiclass Perceptron} algorithm has low time and memory
complexity.

@@

If the examples are strongly linearly separable, there is a simple and efficient
algorithm for the bandit feedback setting that makes at most $?/\gamma^2$
mistakes. The algorithm runs $K$ copies of the \textsc{Binary Perceptron}
algorithm.

For the standard notion of linear separability, we propose three algorithms with
various time-complexity-versus-mistake-bound trade-offs. Two of these algorithms
map the examples into a high-dimensional vector space induced by a kernel. In
this space the examples become \emph{strongly linearly separable}. This allows
us to run a kernelized version of the algorithm for the strongly linearly
separable case.

@@

The examples are said to be \emph{linearly separable with a
margin $\gamma$} if there exist vectors $w_1, w_2, \dots, w_K \in V$ such
that
\begin{align}
\label{equation:linear-separability-1}
& \sum_{i=1}^K \norm{w_i}^2 \le 1 \; , \\
\label{equation:linear-separability-2}
& \forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \qquad \qquad \ip{x_t}{w_{y_t}} \ge \ip{x_t}{w_i} + \gamma \; .
\end{align}

@@

It is easy to see that if labeled examples are strongly linearly
separable with margin $\gamma$ then they are linearly separable with margin
$2\gamma/\sqrt{K}$. Indeed, if $w_1, w_2, \dots, w_K \in V$
satisfy \eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} then $w_1' = w_1/\sqrt{K}, w_2' =
w_2/\sqrt{K}, \dots, w_K' = w_K/\sqrt{K}$ satisfy
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} with margin $\gamma' = 2\gamma/\sqrt{K}$.

Similarly, if $K=2$ and labeled examples are linearly separable with a margin
$\gamma$ then they are strongly linearly separable with the same margin
$\gamma$. Indeed, if $w_1, w_2$ satisfy \eqref{equation:linear-separability-1}
and \eqref{equation:linear-separability-2} then $w_1' = w_1 - w_2$ and $w_2' =
w_2 - w_1$ satisfy \eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3}.

However, if $K \ge 3$ there exists a set of labeled examples in an inner product
space that is linearly separable with a positive margin $\gamma$ but it is not
strongly linearly separable for any positive $\gamma$. Such set of examples is
shown in Figure~\ref{figure:linearly-separable-examples-with-margin}.

\begin{figure}
\begin{center}
\input{figures/linearly-separable-examples-with-margin}
\end{center}
\caption{The figure shows labeled examples in $\R^2$. There are $K=3$ classes. The
examples of each class are colored white, gray and black respectively. All
examples lie in the ball of radius $R$ centered at the origin. While the
examples are linearly separable with a positive margin $\gamma$, they are
\emph{not} strongly linearly separable for any positive margin $\gamma$,
however.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

@@

\begin{algorithm}[h]
\caption{Algorithm for strongly linearly separable examples
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots$}
\STATE{Receive feature vector $x_t$}
\STATE{Compute $S_t = \{ i ~:~ \ip{w_i^{(t)}}{x_t} > 0 \}$}
\IF{$|S_t| = 1$}
\STATE{Predict $\widehat y_t$ where $\widehat y_t \in S_t$}
\ELSE
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\ENDIF
\STATE{Receive feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t = y_t}$}
\STATE{Update $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\IF{$z_t = 0$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)}$}
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

>>>>

% Math symbols and commands
\newcommand{\R}{\mathbb{R}}  % set of real numbers
\newcommand{\indicator}[1]{\mathbf{1}\left[#1 \right]} % indicator
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle} % inner product
\newcommand{\norm}[1]{\left\| #1 \right\|}  % norm of a vector or a matrix
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Exp}{\mathbf{E}}  % expected value

@@

We study the \textsc{Online Multiclass Classification with Bandit Feedback}
problem. The protocol of the problem is stated below. The goal is to design an
algorithm that makes small number of mistakes $\sum_{t=1}^T z_t = \sum_{t=1}^T
\indicator{\widehat y_t \neq y_t}$. The problem was initially proposed by
\cite{Kakade-Shalev-Shwartz-Tewari-2008}. The corresponding problem with full
information feedback is solved by the classical \textsc{Multiclass Perceptron}
algorithm and its variants (see e.g.
\cite{Crammer-Dekel-Keshet-Shalev-Shwartz-Singer-2006}). The problem is quite
natural and potential applications include online advertising and online content
optimization where the feedback corresponds to a user's click.

\begin{algorithm}[h]
\caption{\textsc{Online Multiclass Classification with Bandit Feedback}
\label{algorithm:game-protocol}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Environment \emph{secretly} chooses correct class label $y_t \in \{1,2,\dots,K\}$}
\STATE{Predict class label $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \begin{cases}
0 & \text{if $\widehat y_t = y_t$,} \\
1 & \text{if $\widehat y_t \neq y_t$.}
\end{cases}$
}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

In this paper, we focus on the special case when the examples are \emph{linearly
separable with a margin}. In the full-information feedback setting, it is well
know that if the examples have norm at most $R$ and are linearly separable with
a margin $\gamma$ then the \textsc{Multiclass Perceptron} algorithm makes at
most $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes. This result is very satisfactory
since the upper bound on the number of mistakes is information-theoretically
optimal and at the same time the \textsc{Multiclass Perceptron} algorithm has
low time and memory complexity.

@@

If the examples are strongly linearly separable, there is a simple and efficient
algorithm for the bandit feedback setting with expected number of mistakes
at most $K (R/\gamma)^2$. The algorithm can be viewed as running $K$ copies of
the \textsc{Binary Perceptron} algorithm, one copy for each class.

For the standard notion of linear separability, we propose three algorithms with
various time-complexity-versus-mistake-bound trade-offs. Two of these algorithms
map the examples into a high-dimensional vector space induced by a kernel. In
this space the examples become \emph{strongly} linearly separable. This allows
us to run a kernelized version of the algorithm for the strongly linearly
separable case.

@@

The examples are said to be \emph{linearly separable with a
margin $\gamma$} if there exist vectors $w_1, w_2, \dots, w_K \in V$ such
that
\begin{align}
\label{equation:linear-separability-1}
\sum_{i=1}^K \norm{w_i}^2 & \le 1 \; , \\
\label{equation:linear-separability-2}
\forall t \in \{1,2,\dots,T\} \quad \forall i \in \{1,2,\dots, K\} \setminus \{y_t\} \qquad \qquad \ip{x_t}{w_{y_t}} & \ge \ip{x_t}{w_i} + \gamma \; .
\end{align}

@@

It is easy to see that if labeled examples are strongly linearly
separable with margin $\gamma$ then they are linearly separable with margin
$2\gamma$. Indeed, if $w_1, w_2, \dots, w_K \in V$
satisfy \eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} then they satisfy
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} with margin $\gamma' = 2\gamma$.

Similarly, if $K=2$ and labeled examples are linearly separable with a margin
$\gamma$ then they are strongly linearly separable with margin $\gamma/2$.
Indeed, if $w_1, w_2$ satisfy \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2} then $w_1' = \frac{w_1 - w_2}{2}$,
$w_2' = \frac{w_2 - w_1}{2}$ and $\gamma'=\gamma/2$ satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2},
\eqref{equation:strong-linear-separability-3} since $\norm{w_i'}^2 \le
(\frac{1}{2} \norm{w_1} + \frac{1}{2} \norm{w_2})^2 \le \frac{1}{2}\norm{w_1}^2 +
\frac{1}{2}\norm{w_2}^2$ for $i=1,2$.

However, for $K \ge 3$ there exists a set of labeled examples in an inner product
space that is linearly separable with a positive margin $\gamma$ but it is not
strongly linearly separable with any positive margin $\gamma$. Such set of labeled
examples is shown in
Figure~\ref{figure:linearly-separable-examples-with-margin}.

\begin{figure}
\begin{center}
\input{figures/linearly-separable-examples-with-margin}
\end{center}
\caption{The figure shows labeled examples in $\R^2$. The examples belong to
$K=3$ classes colored white, gray and black respectively. Each class lies in a
$120^\circ$ wedge. While the examples are linearly separable with a positive
margin $\gamma$, they are \emph{not} strongly linearly separable for any
positive margin $\gamma$.
}
\label{figure:linearly-separable-examples-with-margin}
\end{figure}

The notion of linear separability with a margin is standard. It used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm below. It is a folklore result that if a set of labeled examples is
separable with a margin $\gamma$ and the norm of the examples is bounded by $R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\left\lfloor
(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore result is that
\textsc{Multiclass Perceptron} is an optimal algorithm in the sense that any
deterministic algorithm makes $\left\lfloor (R/\gamma)^2 \right \rfloor$
mistakes in the worst case. For completeness, we give proofs  in
Section~\ref{section:multiclass-perceptron-proofs} for both of these results.

\begin{algorithm}[h]
\caption{\textsc{Multiclass Perceptron}
\label{algorithm:mutliclass-perceptron}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t = \argmax_{i \in \{1,2,\dots,K\}} \ip{w_t^{(i)}}{x_t}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\IF{$\widehat y_t \neq y_t$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}$}
\STATE{Update $w_{y_t}^{(t+1)} = w_{y_t}^{(t)} + x_t$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[h]
\caption{Algorithm for strongly linearly separable examples
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Receive feature vector $x_t$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\ELSE
\STATE{Predict $\widehat y_t \sim \text{Uniform}(S_t)$}
\ENDIF
\STATE{Receive feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$S_t = \emptyset$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\ENDIF
\ELSE
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\appendix

\section{Multiclass Perceptron}
\label{section:multiclass-perceptron-proofs}

\begin{theorem}[Upper bound]
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $M = \sum_{t=1}^T \indicator{\widehat y_t \neq y_t}$ be the number of
mistakes the algorithm makes. Since the $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots,
w_K^{(t)})$ changes only if a mistake is made, we can upper bound $\sum_{i=1}^K
\norm{w_i^{(t)}}^2$ in terms of number of mistakes.
If a mistake happens in round $t$,
\begin{align*}
\sum_{i=1}^K \norm{w_i^{(t+1)}}^2
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)} + x_t}^2 + \norm{w_{\widehat y_t}^{(t)} - x_t}^2 \\
& = \left(\sum_{i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\} } \norm{w_i^{(t)}}^2 \right) + \norm{w_{y_t}^{(t)}}^2 + \norm{w_{\widehat y_t}^{(t)}}^2 + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& = \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 + 2 \ip{w_{y_t}^{(t)} - w_{\widehat y_t}^{(t)}}{x_t} \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 \norm{x_t}^2 \\
& \le \left(\sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + 2 R^2 \; .
\end{align*}
So each time a mistake happens, $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ increases by at most $2R^2$. Thus,
$$
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors satisfying the
\eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We lower bound $\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$. This quantity changes
only when a mistakes happens. If mistake happens in round $t$, we have
\begin{align*}
\sum_{i=1}^K \ip{w_i^*}{w_i^{(t+1)}}
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus {y_t, \widehat y_t}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^*}{w_{y_t}^{(t)} + x_t} + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} - x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{y_t}^* - w_{\widehat y_t}^*}{x_t} \\
& \ge  \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \gamma \; .
\end{align*}
Thus, after $M$ mistakes,
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}} \ge \gamma M \; .
$$
We upper bound the left hand side by using Cauchy-Schwartz inequality twice and
the condition \eqref{equation:linear-separability-1} on $w_1^*, w_2^*, \dots,
w_K^*$. We have
$$
\sum_{i=1}^K \ip{w_i^*}{w_i^{(T+1)}}
\le \sum_{i=1}^K \norm{w_i^*} \cdot \norm{w_i^{(T+1)}}
\le \sqrt{\sum_{i=1}^K \norm{w_i^*}^2} \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2}
\le \sqrt{\sum_{i=1}^K \norm{w_i^{(T+1)}}^2} \; .
$$
Combining all inequalities, we get
$$
(\gamma M)^2 \le \sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le 2R^2 M \; .
$$
We conclude that $M \le 2(R/\gamma)^2$. Since $M$ is an integer, $M \le \lfloor 2(R/\gamma)^2 \rfloor$.
\end{proof}

\begin{theorem}[Lower bound]
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
full-information multiclass classification there exists an inner product space
$(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ examples in $V \times
\{1,2,\dots,K\}$ that are linearly separable with margin $\gamma$, the norms
satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the algorithm
makes at least $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}
